---
title: "MLB Attendance Prediction"
description: "Predictive modeling project analyzing how MLB team statistics impact home game attendance using linear regression and Lasso in R"
output: 
  html_document:
    keep_md: true
editor_options: 
  markdown: 
    wrap: 72
---

**Question**: Can we predict an MLB team's attendance based on their
team statistics and player performance?

**Modeling Objective**: Create a predictive model for home attendance
using team statistics from the MLB dataset.

The dataset that I chose (mlb_teams.csv) includes multiple team
statistics across MLB seasons, including wins, runs, home runs, ERA,
walks, stolen bases, and more.
<https://www.openintro.org/data/index.php?data=mlb_teams>

**Response Variable (Y)**: home_attendance – total home game attendance
for the team in a season.

**Predictor Variables (X)** : team statistic variables such as wins,
runs scored, homeruns, strikeouts, shutouts, and more.

Imported the necessary libraries and removed all NA values and
associated rows.

```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(tidymodels)
library(glmnet)

tidymodels::tidymodels_prefer() 

mlbdata <- read.csv("mlb_teams.csv")

mlb <- na.omit(mlbdata)
```

**Exploratory Data Analysis**:

Shown from the summary, there is a wide range of fans attending games
because this data dates back to the late 1800s where attendance was
heavily influenced by stadium capacity, the limited amount of
technology, the mainstream rise of baseball, and historic events such as
wars and economic depressions. This caused the early years of baseball
to smaller amounts of fans in attendance during games. Over time,
attendance trends shifted dramatically due to multiple reasons: better
stadiums were built that could hold more people, expansion of teams in
the league, and improvements to media coverage.

Another trend that is apparent is the positive relationship between team
performance and attendance. Teams that recorded a higher amount of wins,
homeruns, and runs scored saw greater fan turnout. This suggests that
fans are more likely to attend games when their teams are statistically
doing good. However, the scatter plots show a lot of variance,
indicating that these statistics don't fully explain home attendance.
This is also shown through the R-square values of each of the linear
models. Each of the linear regression models show a low amount of
R-squared values, ranging in the low 20%s, indicating that the model can
only explain a small amount of the variance in attendance. This
suggestions that other variables are a big role in attendance.

I also created a histogram of home attendance variable to see the
distribution of the values. The histogram showed a slightly right-skewed
distribution, with most teams hovering around the 2,000,000 value. This
shows us that while attendance is usually consistent, some teams draw
more fans while some draw less fans.

```{r}
summary(mlb$home_attendance)
```

```{r}
options(scipen = 999)

# Linear regression of Wins vs home attendance
ggplot(mlb, aes(x = wins, y = home_attendance)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(title = "Wins vs Attendance")

model1 <- lm(home_attendance ~ wins, data = mlb)
summary(model1)
```

```{r}
# Linear regression Runs scored vs home attendance
ggplot(mlb, aes(x = runs_scored, y = home_attendance)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(title = "Runs Scored vs Attendance")

model2 <- lm(home_attendance ~ runs_scored, data = mlb)
summary(model2)
```

```{r}
# Linear regression of Homeruns vs home attendance
ggplot(mlb, aes(x = homeruns, y = home_attendance)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(title = "Earned Run Average vs Attendance")

model3 <- lm(home_attendance ~ homeruns, data = mlb)
summary(model3)
```

```{r}
# Histogram of home attendance
ggplot(mlb, aes(x = home_attendance)) +
  geom_histogram(binwidth = 500000, color = "white") +
  labs(title = "Distribution of Home Attendance",
       x = "Home Attendance",
       y = "Frequency") +
  theme_minimal()
```

**Feature Engineering:**

I created two new variables to better measure team performance:

1.  win percentage

2.  runs per game

These features help compare teams more fairly across different seasons
and makes the model more accurate.

```{r}
mlb$win_pct = mlb$wins / (mlb$wins+mlb$losses)
mlb$runs_per_game = mlb$runs_scored/mlb$games_played
```

**Model Selection:**

I chose linear regression as my primary model since the outcome, home
attendance, is continuous, making linear regression the most natural
starting point. Linear regression allows us to understand the direction
and the strength of the relationships between team statistics and home
attendance easily.

**Model Training and Evaluation:**

To find how effective the linear regression model was at predicting, I
split the dataset into two categories: training and test sets, with the
training set having 80% of the data while the test set having 20% of the
data. I then put the variables **wins**, **homeruns**, **runs scored**,
and **hits** as predictor variables and normalized them. After finding
the predicted values using augment(), I found the RMSE, MAE, and
R-squared valued it showed that the model wasn't highly accurate, with
the RMSE and MAE value fairly high with the R-squared value fairly low.
I also graphed the model to see the the predicted values versus the
actual values, and it showed that there was still a high variance.

```{r}
set.seed(1)

mlb_split <- mlb %>% initial_split(prop = 0.80)
mlb_train <- training(mlb_split)
mlb_test <- testing(mlb_split)

mlb_recipe <- recipe(home_attendance ~ wins + homeruns + runs_scored + hits, data = mlb_train) %>%
  step_normalize(all_predictors())

model <- linear_reg()

mlb_wflow <- workflow() %>%
  add_model(model) %>%
  add_recipe(mlb_recipe)

mlb_fit <- fit(mlb_wflow, data = mlb_train)

results <- augment(mlb_fit, new_data = mlb_test)

results %>% # high variance shown within the graph
  ggplot(aes(x = home_attendance, y = .pred)) +
  geom_abline(lty = 2, color="red") +
  geom_point(alpha = 0.7, color="blue") +
  labs(title = "Predicted versus actual values of home attendance", 
       x = "Home Attendance", 
       y = "Predicted Home Attendance") +
  coord_obs_pred()

mlb_metrics <- metric_set(rmse, rsq, mae)
mlb_metrics(results, truth = home_attendance, estimate = .pred)
```

**Model Tuning:**

I used Lasso Regression and Cross-Validation to identify the most
important predictors of home attendance. Lasso regression applies an L1
penalty, shrinking the less important coefficients towards zero,
sometimes setting them to exactly zero. The results showed that the
strongest predictors included **wins**, **runs_per_game**, **hits**,
**homeruns**, and **strikeouts_by_pitcher**. This allowed me to tune the
parameters for the final model.

```{r}
set.seed(1)

X <- model.matrix(home_attendance ~ wins + homeruns + runs_scored + hits + strikeouts_by_pitchers + doubles + triples + walks + stolen_bases + complete_games + shutouts + saves + double_plays + earned_run_average + win_pct + runs_per_game, data = mlb)[, -1] 
y <- mlb$home_attendance

cv_model <- cv.glmnet(X, y, alpha = 1)

best_lambda <- cv_model$lambda.min
best_lambda

X_scaled <- scale(X)
lasso_model <- glmnet(X_scaled, y, alpha = 1, lambda = best_lambda, standardize = TRUE)

coef(lasso_model)
```

**Final Model**:

The final model assumes that the relationship between the team
statistics and home attendance is linear and that the points are
independent of each other.

```{r}
options(scipen = 999)

set.seed(1)
mlb_split <- mlb %>% initial_split(prop = 0.80)
mlb_train <- training(mlb_split)
mlb_test <- testing(mlb_split)

mlb_recipe <- recipe(home_attendance ~ wins + runs_per_game + hits + homeruns + strikeouts_by_pitchers, data = mlb_train) %>%     step_normalize(all_predictors())

model <- linear_reg() 

mlb_wflow <- workflow() %>%
  add_model(model) %>%
  add_recipe(mlb_recipe)

mlb_fit <- fit(mlb_wflow, data = mlb_train)

results <- augment(mlb_fit, new_data = mlb_test)


results %>% # high variance shown within the graph
  ggplot(aes(x = home_attendance, y = .pred)) +
  geom_abline(lty = 2, color="red") +
  geom_point(alpha = 0.7, color="blue") +
  labs(title = "Final Model: Predicted versus actual values of \nhome attendance", 
       x = "Home Attendance", 
       y = "Predicted Home Attendance") +
  coord_obs_pred()

mlb_metrics <- metric_set(rmse, rsq, mae)
mlb_metrics(results, truth = home_attendance, estimate = .pred)
```

**Interpretations**:

Overall, the model's performance was still not strong enough to be
considered highly predictive. This indicates that while team statistics
can help explain some of the variation in home attendance, other
external factors such as weather, star players, promotions, market size,
and rivalry games likely play a big role in home attendance.
